---
title: "Fit sparse satellite data with inlabru"
author: "Devaraj Gopinathan"
date: "Generated on `r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    df_print: paged
---

### Importing libraries

```{r results = "hide", warning = FALSE, message = FALSE}
library(INLA)
library(inlabru)
library(fmesher)
library(ggplot2)
library(patchwork)
library(terra)
library(dplyr)
library(tidyr)
```

### Importing data

```{r results = "hide", warning = FALSE, message = FALSE}
data <- readRDS("../outputs/sst.rds")
var <- 'SST'
limits <- c(range(data['x'])[1], range(data['x'])[2], range(data['y'])[1], range(data['y'])[2])
```

### Create colour scale
```{r}
colsc <- function(...) {
  scale_fill_gradientn(
    colours = rev(RColorBrewer::brewer.pal(11, "RdYlBu")),
    name = 'SST',
    limits = range(..., na.rm = TRUE)
  )
}
```

### Remove na values from observations
```{r warning=FALSE}
plot <- ggplot(data) +
  geom_raster(aes(x = x, y = y, fill = SST)) +
  scale_fill_gradientn(colors = terrain.colors(100)) +
  coord_equal()

(plot)

print(paste0("obs: no. of total points is ", ncell(data['SST'])))
print(paste0("obs: no. of NAs is ", sum(is.na(data['SST']))))
print(paste0("obs: no. of valid data points is ", sum(!is.na(data['SST']))))

data = data %>% drop_na(var)

plot <- ggplot(data) +
  geom_raster(aes(x = x, y = y, fill = SST)) +
  scale_fill_gradientn(colors = terrain.colors(100)) +
  coord_equal()

(plot)

print(paste0("obs: no. of total points is ", ncell(data['SST'])))
print(paste0("obs: no. of NAs is ", sum(is.na(data['SST']))))
print(paste0("obs: no. of valid data points is ", sum(!is.na(data['SST']))))

```
### Sample observations
```{r}
set.seed(1931)
sample.n <- nrow(data) # 1000
sample.index = sample(1:nrow(data), sample.n)
sample.data = data[sample.index,]
sample_df <- sf::st_as_sf(
  data.frame(x = sample.data['x'], y = sample.data['y']),
  coords = c("x", "y")
)
sample_df['SST'] <- sample.data['SST']

plot <- ggplot() +
  gg(sample_df, aes(colour = SST)) +
  scale_color_gradientn(colors = terrain.colors(100)) +
  coord_sf()

(plot)
```
### Pre-requisites to fit random field - 1 - Create mesh
```{r}
region_of_interest <- spoly(
  data.frame(
    x = limits[c(1, 2, 2, 1)],
    y = limits[c(3, 3, 4, 4)]
  ),
  format = "sf"
)

mesh <- fm_mesh_2d(
  loc = fm_hexagon_lattice(sample_df, edge_len = 0.2),
  boundary = region_of_interest,
  max.edge = 0.3
)

plot <- ggplot() +
  geom_fm(data = mesh) +
  gg(sample_df, aes(colour = SST)) +
  scale_color_gradientn(colors = terrain.colors(100)) +
  coord_sf()

(plot)
```
### Pre-requisites to fit random field - 2 - Specify spde model
```{r}
matern <-
  inla.spde2.pcmatern(mesh,
    prior.sigma = c(10, 0.01),
    prior.range = c(1, 0.01)
  )
```

### Pre-requisites to fit random field - 3 - Specify model components
```{r}
cmp <- SST ~ field(geometry, model = matern) + Intercept(1)
```

### Fit random field
```{r}
fit <- bru(cmp, sample_df, family = "gaussian")
summary(fit)
```
### Predict field on a grid
```{r}
grid <- fm_pixels(mesh, dims = c(200, 200))
pred <- predict(
  fit,
  grid,
  ~ field + Intercept
)
```

### Generate samples
```{r}
n_samples = 100
sample <- generate(
  fit,
  grid,
  ~ field + Intercept,
  n.samples = n_samples
)
pred$sample <- sample
```

### Plot mean and sd of predicted field
```{r warning=FALSE}
# Common colour scale for the truth and estimate:
csc <- colsc(data$SST, pred$mean, pred$sample)

plot_sst <- ggplot(data) +
  geom_raster(aes(x = x, y = y, fill = SST)) +
  gg(region_of_interest, alpha = 0) + csc +
  ggtitle("SST") 

plot_posterior_mean <- ggplot() +
  gg(pred, aes(fill = mean), geom = "tile") +
  gg(region_of_interest, alpha = 0) + csc +
  ggtitle("Posterior mean")

plot_posterior_sd <- ggplot() +
  gg(pred, aes(color = sd)) +
  gg(region_of_interest, alpha = 0) +
  scale_colour_viridis_c(direction = -1, name = 'SST') +
  ggtitle("Posterior sd")

(plot_sst)
(plot_posterior_mean)
(plot_posterior_sd)
```
### Store samples
```{r}
n_sample_plots = 7
plot_posterior_sample <- vector("list", n_sample_plots)
count = 1
for (i in sample(1:n_samples, 5)) {
  
pred$temp <- pred$sample[, i]
plot_posterior_sample[[count]] <- ggplot() +
  gg(pred, aes(fill = temp), geom = "tile") +
  gg(region_of_interest, alpha = 0) + csc +
  ggtitle(paste0("Posterior sample ", i)) 
count <- count + 1
}
```

### Plot samples
```{r warning=FALSE}
(plot_posterior_sample[[1]])
(plot_posterior_sample[[2]])
(plot_posterior_sample[[3]])
(plot_posterior_sample[[4]])
(plot_posterior_sample[[5]])
```
### Write samples to csv
```{r}
sample_xy <- sf::st_coordinates(grid)
sample_df <- data.frame(
  x = sample_xy[,1],
  y = sample_xy[,2],
  value = pred$sample
)
sample_df[, 3:ncol(sample_df)] <- lapply(sample_df[, 3:ncol(sample_df)], round, digits = 4)
write.csv(sample_df, "../outputs/samples.csv", row.names = FALSE)
```

### Read and plot from file for verification
```{r}
temp <- read.csv("../outputs/samples.csv")

i = sample(1:n_samples, 1)
plot_sample_file <- ggplot(temp) +
  geom_raster(aes(x = x, y = y, fill = .data[[paste0("value.", i)]])) +
  gg(region_of_interest, alpha = 0) + csc +
  ggtitle(paste0("Posterior sample ", i)) 

pred$temp <- pred$sample[, i]
plot_sample_calc <- ggplot() +
  gg(pred, aes(fill = temp), geom = "tile") +
  gg(region_of_interest, alpha = 0) + csc +
  ggtitle(paste0("Posterior sample ", i)) 

(plot_sample_file)
(plot_sample_calc)
```

Acknowledgment: based on `inla-bru` vignette at *https://inlabru-org.github.io/inlabru/articles/random_fields_2d.html*
